The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-04-14 paper/iphas-dr2.tex
Changes based on feedback by SG, JE and DS.

\def\ha{\mbox{H$\rm \alpha$}}
\def\arcsec{$''$}
\def\arcmin{$'$}
\def\deg{$^{\circ}$}
\def\micron{\mbox{$\mu$m}}

@@

\author[G. Barentsen
et. al]{G. Barentsen$^{1}$\thanks{E-mail:geert@barentsen.be},
H. J. Farnhill$^1$,
J. E. Drew$^1$,
E. A. Gonz$\acute{\rm{a}}$lez-Solares$^2$, 
R. Greimel$^3$, \newauthor
M. J. Irwin$^2$,
B. Mizalski$^{4}$,
C. Ruhland$^1$,
P. Groot$^5$,
A. Mampaso$^6$, 
S. E. Sale$^7$, \newauthor
M. J. Barlow$^{8}$,
R. L. M Corradi$^{6}$,
J. J. Drake$^{9}$,
J. Eisl\"offel$^{10}$,
J. Fabregat$^{11}$, \newauthor
B. T. Gaensicke$^{12}$,
A. S. Hales$^{13}$,
J. Irwin$^{9}$,
C. Knigge$^{14}$,
T. Kupfer$^{5}$,
D. J. Lennon$^{15}$, \newauthor
J. R. Lewis$^{2}$, 
M. Mohr-Smith$^{1}$, 
R. A. H. Morris$^{16}$,
T. Naylor$^{17}$,
Q. A. Parker$^{18}$, \newauthor
S. Phillipps$^{13}$, 
R. Raddi$^{11}$, 
P. Rodriguez-Gil$^{6}$,
L. Sabin$^{19}$,
S. Scaringi$^{20}$,
D. Steeghs$^{11}$, \newauthor
Y. C. Unruh$^{21}$, 
K. Viironen$^{22}$, 
J. S. Vink$^{23}$, 
N. A. Walton$^{2}$,
N. J. Wright$^{1}$,
A. A. Zijlstra$^{24}$.
\\
$^{1}$School of Physics, Astronomy \& Mathematics, University of Hertfordshire, College Lane, Hatfield, Hertfordshire, AL10 9AB, U.K.\\
$^{2}$Institute of Astronomy, University of Cambridge, Madingley Road, Cambridge, CB3 OHA, U.K.\\
$^{3}$IGAM, Institute of Physics, University of Graz, Universit\"atsplatz 5, Graz, Austria.\\
$^{4}$South African Astronomical Observatory, P.O. Box 9, Observatory, 7935 Cape Town, South Africa.\\
$^{5}$Afdeling Sterrenkunde, Radboud Universiteit Nijmegen, Faculteit NWI, Postbus 9010, 6500 GL Nijmegen, The Netherlands.\\
$^{6}$Instituto de Astrof\'isica de Canarias, 38200 La Laguna, Tenerife, Spain.\\
$^{7}$Rudolf Peierls Centre for Theoretical Physics, Keble Road, Oxford, OX1 3NP, U.K.\\
$^{8}$University College London, Department of Physics \& Astronomy, 
Gower Street, London WC1E 6BT, U.K.\\
$^{9}$Harvard-Smithsonian Center for Astrophysics, 60 Garden Street, 
Cambridge, MA 02138, U.S.A. \\
$^{10}$Th\"uringer Landessternwarte, Sternwarte 5, 07778, Tautenburg, Germany \\
$^{11}$Observatorio Astr\'onomico, Universidad de Valencia,
Catedr\'atico Jos\'e Beltr\'an 2, 46980 Paterna, Spain\\
$^{12}$Department of Physics, University of Warwick, Gibbet Hill Road, Coventry, CV4 7AL, U.K.\\
$^{13}$Joint ALMA Observatory, Alonso de Córdova 3107, Vitacura 763-0355, Santiago, Chile. \\
$^{14}$School of Physics \& Astronomy, University of Southampton, Southampton, SO17 1BJ, U.K.\\
$^{15}$European Space Astronomy Centre (ESAC), Villafranca del Castillo, Villanueva de la Canada, E-28692 Madrid, Spain.\\
$^{16}$School of Physics, Bristol University, Tyndall Avenue, Bristol, BS8 1TL, U.K.\\
$^{17}$School of Physics, University of Exeter, Stocker Road, Exeter, EX4 4QL, U.K.\\
$^{18}$Department of Physics \& Astronomy, Macquarie University, NSW 2109, Australia\\
$^{19}$Instituto de Astonom\'ia y Meteorolog\'ia, Departamento de F\'isica, CUCEI, Universidad de Guadalajara, Mexico.\\
$^{20}$Instituut voor Sterrenkunde, K.U. Leuven, Celestijnenlaan 200D, B-3001 Leuven, Belgium\\
$^{21}$Department of Physics, Blackett Laboratory, Imperial College
London, Prince Consort Road, London, SW7 2AZ, U.K.\\
$^{22}$Centro de Estudios de F\'isica del Cosmos de Arag\'on, Plaza San Juan 1, Planta 2, Teruel, 44001, Spain. \\
$^{23}$Armagh Observatory, College Hill, Armagh, Northern Ireland,
BT61 9DG, U.K.\\
$^{24}$Jodrell Bank Centre for Astrophysics, School of Physics \&
Astronomy, University of Manchester, Manchester M13 9PL, U.K.}

@@

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800~deg$^2$ imaging survey
covering Galactic latitudes $|b| < 5^\circ$
and longitudes $l=30$ to $215^\circ$
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
uniformly-calibrated source catalogue
derived from the survey,
providing single-epoch photometry
for 219 million unique sources
across 92\% of the footprint.
The observations were carried out between 2003 and 2013
at a median seeing of 1.1 arcsec
and to a mean $5\sigma$-depth of 
21.2 ($r$), 20.0 ($i$) and 20.3 ($\ha$)
in the Vega magnitude system.
We explain the data reduction 
and quality control procedures,
describe and test the new uniform calibration,
and detail the construction of the new catalogue.
We show that the global calibration is accurate to
0.03~mag (rms) 
and recommend a series of quality criteria
to select the most reliable data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-\ha,\ r-i)$ diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sight-lines
and (ii) select candidate \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to recent infrared surveys.
\end{abstract}

@@

\section{Introduction}
The INT/WFC Photometric H$\alpha$ Survey
of the Northern Galactic Plane \citep[IPHAS;][]{Drew2005}
is providing new insights into the contents and structure of the disk of the Milky Way.
This large-scale programme of observation
-- spanning almost a decade 
and using more than 300 nights 
at the Isaac Newton Telescope (INT) in La Palma --
aims to provide the digital update 
to the photographic northern H$\alpha$ surveys 
of the mid-20th century \citep[see][]{Kohoutek1999}. 
By increasing the sensitivity 
with respect to these previous surveys 
by a factor $\sim$1000 (7 magnitudes), 
IPHAS can expand
the limited bright samples of Galactic emission line objects 
previously available into larger, deeper, and more statistically-robust samples that will 
better inform our understanding 
of the early and late stages of stellar evolution.
Since the publication of the IPHAS Initial Data Release \citep[IDR;][]{Gonzalez-Solares2008},
these aims have begun to be realised through a
range of published studies including: 
a preliminary catalogue of candidate emission-line objects \citep{Witham2008};
discoveries of new symbiotic stars \citep{Corradi2008, Corradi2010, Corradi2011}; 
new cataclysmic variables \citep{Witham2007}; 
new groups of young stellar objects
\citep{Vink2008,Barentsen2011a,Wright2012};
new classical Be stars \citep{Raddi2013};
along with discoveries of new supernova remnants \citep{Sabin2013}
and new and remarkable planetary nebulae 
\citep{Mampaso2006, Viironen2009a, Viironen2009b, Sabin2010, Corradi2011, Viironen2011}.

Over the years it has become apparent that the legacy of IPHAS 
will reach beyond these traditional \ha\ applications 
of identifying emission-line stars and nebulae. 
Through the provision of $r$, $i$ broadband photometry 
alongside H$\alpha$ data,
IPHAS has created the opportunity 
to study Galactic Plane populations 
in a new way.
For example, the survey’s unique $(r-\ha,\ r-i)$ colour-colour
diagram has been shown to provide simultaneous constraints 
on intrinsic stellar colour and interstellar extinction \citep{Drew2008}. 
This has opened the door 
to a wide range of Galactic science applications, 
including the mapping of extinction across the Plane in three dimensions
and the probabilistic inference of stellar properties
\citep{Sale2009, Sale2010, Giammanco2011, Sale2012, Barentsen2013}. 
In effect, the availability of narrowband H$\alpha$ 
alongside $r, i$ magnitudes
provides coarse spectral information for huge samples of stars 
which are otherwise too faint or numerous 
to be targeted by spectroscopic surveys.
For such science applications to succeed, however, 
it is vital that the imaging data are transformed 
into a uniformly-calibrated photometric catalogue, 
in which quality problems 
and duplicate detections are flagged.

When the initial data release was created in late 2007,
just over $\sim$half of the survey footprint was covered
and the data were insufficiently complete 
to support a homogeneously calibrated source catalogue.
The goal of this paper is to present the next release 
that takes the coverage up to 92 percent of the survey area 
and includes a uniform calibration.
In this work we
(i) explain the data reduction 
and quality control procedures that were applied,
(ii) describe and test the new global photometric calibration, and 
(iii) detail the construction of the source catalogue
and demonstrate its use.

In \S\ref{sec:observations} we start by recapitulating the key points
of the survey observing strategy.
In \S\ref{sec:reduction} we describe the data reduction
and quality control procedures.
In \S\ref{sec:calibration} we explain the uniform re-calibration, in which
we draw upon the AAVSO Photometric All-Sky Survey (APASS)
and test our results against the Sloan Digital Sky Survey (SDSS).
In \S\ref{sec:catalogue} we explain how the source catalogue was compiled.
In \S\ref{sec:discussion} we discuss the properties of the catalogue
and in \S\ref{sec:demonstration} we demonstrate
the scientific exploitation of the colour/magnitude diagrams.
Finally, in \S\ref{sec:dataaccess} we discuss access
to the catalogue, an online library of reduced images and relevant source code,
The paper ends with conclusions in \S\ref{sec:conclusions} where we also outline
our future ambitions.

@@

The detailed properties of the IPHAS observing programme 
have been presented before 
by \citet{Drew2005} and \citet{Gonzalez-Solares2008}. 
To set the stage for this release, we provide a reminder of some key points.
IPHAS is an imaging survey of the Galactic Plane north of the celestial equator, 
from which photometry in Sloan $r, i$ 
is extracted along with narrowband H$\alpha$. 
It is carried out using the Wide Field Camera (WFC) 
on the 2.5-metre INT in La Palma. 
It is the first digital survey to offer comprehensive optical CCD photometry
of point sources in the Galactic Plane;
the footprint spans a box 
of roughly 180 by 10 degrees, 
covering Galactic latitudes $-5^{\circ} < b < +5^{\circ}$ 
and longitudes $30^{\circ} < l < 215^{\circ}$.

@@

\begin{figure*}
        \includegraphics[width=1\linewidth]{./plots/footprint/footprint_small.png}
        \caption{Survey area showing the footprints
        of all the quality-approved IPHAS fields
        which have been included in this data release.
        The area covered by each field has been coloured black
        with a semi-transparent opacity of 20\%,
        such that regions where fields overlap are darker.
        The IPHAS strategy is to observe each field twice
        with a small offset,
        and hence the vast majority of the area 
        is covered twice (dominant grey colour).
        There are small overlaps between all the neighbouring fields
        which can be seen as a honeycomb pattern
        of dark grey lines across the survey area.
        Regions with incomplete data are apparent as white gaps (no data) 
        or in light grey (denoting that only one offset is missing).
        The dark vertical strip near $l \sim$125\deg
        is an arbitrary consequence of the tiling pattern,
        which was populated starting from 0h in right ascension.}
        \label{fig:footprint}
\end{figure*}

Fig.~\ref{fig:footprint} shows the footprin
of the quality-approved observations included in this work. 
The fields which remain missing 
-- covering 8\% of the survey area --
are predominantly located towards the Galactic anti-centre 
at $l > 120^o$.
Fields at these longitudes are mainly accessed from La Palma 
in the months of November-December,
which is when the La Palma weather and seeing conditions are often poor
forcing many (unsuccessful) repeat observations.
To enable the survey to be brought to completion, 
a decision was made recently to limit repeats in this area 
to individual fields requiring replacement,
i.e. fresh observations in all 3 filters may only be obtained 
at one of the two offset positions.  
The catalogue is structured such that it is clear 
where a contemporaneous observation of both halves of a field pair
is available.

@@

All raw IPHAS data were transferred
to the Cambridge Astronomical Survey Unit (CASU) 
for initial processing and archival.
The procedures used by CASU were originally devised
for the INT Wide Field imaging Survey \citep[WFS;][]{McMahon2001,Irwin2005},
which was a 200 deg$^2$ survey programme carried out 
between 1998 and 2003 after the WFC was commissioned.
Because IPHAS uses the same telescope and camera combination,
we have been able to benefit from the existing WFS pipeline.
A detailed description of the processing steps 
is found in \citet{Irwin2001}.
Its application to IPHAS has previously been described
by \citet{Drew2005} and \citet{Gonzalez-Solares2008},
and much of the source code is available 
on line\footnote{http://casu.ast.cam.ac.uk/surveys-projects/software-release}. 
In brief, the pipeline takes care of bias subtraction,
linearity correction, flat-fielding,
gain correction and de-fringing.

The reduced images are then stored in multi-extension FITS files
with a primary header describing the characteristics
(position, filter, exposure time, etc.) 
and four image extensions 
corresponding to each of the four CCDs.
Source detection and characterisation is then carried out 
using the \textsc{imcore} tool \citep{Irwin1985,Irwin1997}.
The flux of each source is measured using both
the peak pixel height
(i.e. a square 0.33$\arcsec\times$0.33\arcsec\ aperture)
and a series of circular apertures of increasing diameter 
(1.2\arcsec, 2.3\arcsec, 3.3\arcsec, 4.6\arcsec\ and 6.6\arcsec).

@@

Finally, an astrometric solution is determined
based on the Two-Micron All Sky Survey (2MASS) point source catalogue \citep{Skrutskie2006},
which itself is calibrated 
in the International Celestial Reference System (ICRS).
A provisional photometric calibration is also provided 
based on the average zeropoint
determined from sets of standard stars observed within the same night.
Sources are classified morphologically
-- stellar, extended or noise --
based on the curve-of-growth determined
from measuring the source intensity in a series of growing apertures.
Finally, the resulting source detection tables are also stored 
in multi-extension FITS files.

At the time of preparing DR2,
the CASU pipeline had processed
74\,195 IPHAS exposures 
in which a total of 1.9~billion \emph{candidate sources} were detected 
at the sensitive default detection level of 1.25\,$\sigma$.
This total inevitably includes spurious detections, artefacts and
duplicate detections,
in \S\ref{sec:catalogue} we will explain
how these have been removed or flagged in the final catalogue.
The pipelined data set -- comprising 2.5~terabyte of FITS files --
was then transferred to the University of Hertfordshire
for the purpose of transforming the raw
detection tables into a reliable source catalogue which is 
(i) quality-controlled,
(ii) homogeneously calibrated, and 
(iii) contains user-friendly columns and warning flags.
It is these post-processing steps which are explained next.

@@

Observing time for IPHAS was obtained
on a semester-by-semester basis
through the open time allocation committees 
of the Isaac Newton Group of telescopes,
which are invariably over-subscribed.
For this reason, we attempted to utilise 
\emph{all} the nights allocated to IPHAS,
even those which were partially or entirely non-photometric
or otherwise affected by technical problems 
(e.g. electronic noise or telescope tracking problems).
Any unsuitable data that were taken as a result
of this strategy have subsequently been flagged and removed
using seven quality criteria,
which ensure a reliable and homogeneous level of quality
across the data release:

\begin{figure}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=\textwidth]{./plots/depth_r.pdf} 
    \end{minipage}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=\textwidth]{./plots/depth_i.pdf} 
    \end{minipage}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=\textwidth]{./plots/depth_h.pdf} 
    \end{minipage}
    \caption{Distribution of the 5$\sigma$ limiting magnitude
             across all quality-approved survey fields
             for $r$ (top), $i$ (middle) and \ha\ (bottom).
             Fields with a limiting magnitude brighter than
             20th ($r$) or 19th (\ha, $i$) were rejected
             from the data release.
             The $r$-band depth is most sensitive 
             to the presence of the moon above the horizon: 
             this is the reason for the wide, bi-modal character
             of its distribution.}
    \label{fig:depth}
\end{figure}

>>>>

\def\ha{\mbox{H$\rm \alpha$}}
\def\micron{\mbox{$\mu$m}}

@@

\author[G. Barentsen
et. al]{G. Barentsen$^{1}$\thanks{E-mail:geert@barentsen.be},
H. J. Farnhill$^1$,
J. E. Drew$^1$,
E. A. Gonz$\acute{\rm{a}}$lez-Solares$^2$, 
R. Greimel$^3$, \newauthor
M. J. Irwin$^2$,
B. Mizalski$^{4}$,
C. Ruhland$^1$,
P. Groot$^5$,
A. Mampaso$^6$, 
S. E. Sale$^7$, \newauthor
M. J. Barlow$^{8}$,
P. J. Carter$^{12}$,
R. L. M Corradi$^{6}$,
J. J. Drake$^{9}$,
J. Eisl\"offel$^{10}$, \newauthor
J. Fabregat$^{11}$, 
B. T. G\"ansicke$^{12}$,
N. P. Gentile Fusillo$^{12}$,
S. Greiss$^{12}$,
A. S. Hales$^{13}$, \newauthor
S. Hodgkin$^2$, 
J. Irwin$^{9}$,
C. Knigge$^{14}$,
T. Kupfer$^{5}$,
D. J. Lennon$^{15}$,
J. R. Lewis$^{2}$,  \newauthor
M. Mohr-Smith$^{1}$, 
R. A. H. Morris$^{16}$,
T. Naylor$^{17}$,
Q. A. Parker$^{18}$, 
S. Phillipps$^{13}$, \newauthor
R. Raddi$^{11}$, 
P. Rodriguez-Gil$^{6}$,
L. Sabin$^{19}$,
S. Scaringi$^{20}$,
D. Steeghs$^{11}$, 
Y. C. Unruh$^{21}$, \newauthor
J. van Roestel$^5$,
K. Viironen$^{22}$, 
J. S. Vink$^{23}$, 
N. A. Walton$^{2}$,
N. J. Wright$^{1}$,
A. A. Zijlstra$^{24}$.
\\
$^{1}$School of Physics, Astronomy \& Mathematics, University of Hertfordshire, College Lane, Hatfield, Hertfordshire, AL10 9AB, U.K.\\
$^{2}$Institute of Astronomy, University of Cambridge, Madingley Road, Cambridge, CB3 OHA, U.K.\\
$^{3}$IGAM, Institute of Physics, University of Graz, Universit\"atsplatz 5, Graz, Austria.\\
$^{4}$South African Astronomical Observatory, P.O. Box 9, Observatory, 7935 Cape Town, South Africa.\\
$^{5}$Afdeling Sterrenkunde, Radboud Universiteit Nijmegen, Faculteit NWI, Postbus 9010, 6500 GL Nijmegen, The Netherlands.\\
$^{6}$Instituto de Astrof\'isica de Canarias, 38200 La Laguna, Tenerife, Spain.\\
$^{7}$Rudolf Peierls Centre for Theoretical Physics, Keble Road, Oxford, OX1 3NP, U.K.\\
$^{8}$University College London, Department of Physics \& Astronomy, 
Gower Street, London WC1E 6BT, U.K.\\
$^{9}$Harvard-Smithsonian Center for Astrophysics, 60 Garden Street, 
Cambridge, MA 02138, U.S.A. \\
$^{10}$Th\"uringer Landessternwarte, Sternwarte 5, D-07778, Tautenburg, Germany \\
$^{11}$Observatorio Astron\'omico, Universidad de Valencia,
Catedr\'atico Jos\'e Beltr\'an 2, 46980 Paterna, Spain\\
$^{12}$Department of Physics, University of Warwick, Gibbet Hill Road, Coventry, CV4 7AL, U.K.\\
$^{13}$Joint ALMA Observatory, Alonso de Córdova 3107, Vitacura 763-0355, Santiago, Chile. \\
$^{14}$School of Physics \& Astronomy, University of Southampton, Southampton, SO17 1BJ, U.K.\\
$^{15}$European Space Astronomy Centre (ESAC), Villafranca del Castillo, Villanueva de la Canada, E-28692 Madrid, Spain.\\
$^{16}$School of Physics, Bristol University, Tyndall Avenue, Bristol, BS8 1TL, U.K.\\
$^{17}$School of Physics, University of Exeter, Stocker Road, Exeter, EX4 4QL, U.K.\\
$^{18}$Department of Physics \& Astronomy, Macquarie University, NSW 2109, Australia\\
$^{19}$Instituto de Astonom\'ia y Meteorolog\'ia, Departamento de F\'isica, CUCEI, Universidad de Guadalajara, Mexico.\\
$^{20}$Instituut voor Sterrenkunde, K.U. Leuven, Celestijnenlaan 200D, B-3001 Leuven, Belgium\\
$^{21}$Department of Physics, Blackett Laboratory, Imperial College
London, Prince Consort Road, London, SW7 2AZ, U.K.\\
$^{22}$Centro de Estudios de F\'isica del Cosmos de Arag\'on, Plaza San Juan 1, Planta 2, Teruel, 44001, Spain. \\
$^{23}$Armagh Observatory, College Hill, Armagh, Northern Ireland,
BT61 9DG, U.K.\\
$^{24}$Jodrell Bank Centre for Astrophysics, School of Physics \&
Astronomy, University of Manchester, Manchester M13 9PL, U.K.}

@@

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800~deg$^2$ imaging survey
covering Galactic latitudes $|b| < 5\degr$
and longitudes $\ell=30$ to 215\degr
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
globally calibrated source catalogue
derived from the survey,
providing single-epoch photometry
for 219 million unique sources
across 92\% of the footprint.
The observations were carried out between 2003 and 2013
at a median seeing of 1.1~arcsec (sampled at 0.33~arcsec/pixel)
and to a mean $5\sigma$-depth of 
21.2~($r$), 20.0~($i$) and 20.3~($\ha$)
in the Vega magnitude system.
We explain the data reduction 
and quality control procedures,
describe and test the global re-calibration,
and detail the construction of the new catalogue.
We show that the new calibration is accurate to
0.03~mag (rms) 
and recommend a series of quality criteria
to select the most reliable data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-\ha,\ r-i)$ diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sight-lines
and (ii) select \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to recent infrared surveys.
\end{abstract}

@@

\section{Introduction}
The INT/WFC Photometric H$\alpha$ Survey
of the Northern Galactic Plane \citep[IPHAS;][]{Drew2005}
is providing new insights into the contents and structure of the disk of the Milky Way.
This large-scale programme of observation
-- spanning almost a decade 
and using more than 300 nights 
at the Isaac Newton Telescope (INT) in La Palma --
aims to provide the digital update 
to the photographic northern H$\alpha$ surveys 
of the mid-20th century \citep[see][]{Kohoutek1999}. 
By increasing the sensitivity 
with respect to these previous surveys 
by a factor $\sim$1000 (7 magnitudes), 
IPHAS can expand
the limited bright samples of Galactic emission line objects 
previously available into larger, deeper, and more statistically-robust samples that will 
better inform our understanding 
of the early and late stages of stellar evolution.
Since the publication of the Initial Data Release \citep[IDR;][]{Gonzalez-Solares2008},
these aims have begun to be realised through a
range of published studies including: 
a preliminary catalogue of candidate emission-line objects \citep{Witham2008};
discoveries of new symbiotic stars \citep{Corradi2008, Corradi2010, Corradi2011}; 
new cataclysmic variables \citep{Witham2007,Wesson2008,Aungwerojwit2012}; 
new groups of young stellar objects
\citep{Vink2008,Barentsen2011a,Wright2012};
new classical Be stars \citep{Raddi2013};
along with discoveries of new supernova remnants \citep{Sabin2013}
and new and remarkable planetary nebulae 
\citep{Mampaso2006, Viironen2009a, Viironen2009b, Sabin2010, Corradi2011, Viironen2011}.

Over the years it has become apparent that the legacy of IPHAS 
will reach beyond these traditional \ha\ applications 
of identifying emission-line stars and nebulae. 
Through the provision of $r$, $i$ broadband photometry 
alongside H$\alpha$ data,
IPHAS has created the opportunity 
to study Galactic Plane populations 
in a new way.
For example, the survey’s unique $(r-\ha,\ r-i)$ colour-colour
diagram has been shown to provide simultaneous constraints 
on intrinsic stellar colour and interstellar extinction \citep{Drew2008}. 
This has opened the door 
to a wide range of Galactic science applications, 
including the mapping of extinction across the Plane in three dimensions
and the probabilistic inference of stellar properties
\citep{Sale2009, Sale2010, Giammanco2011, Sale2012, Barentsen2013}. 
In effect, the availability of narrowband H$\alpha$ 
alongside $r, i$ magnitudes
provides coarse spectral information for huge samples of stars 
which are otherwise too faint or numerous 
to be targeted by spectroscopic surveys.
For such science applications to succeed, however, 
it is vital that the imaging data are transformed 
into a homogeneously-calibrated photometric catalogue, 
in which quality problems 
and duplicate detections are flagged.

When the initial data release was created in late 2007,
just over $\sim$half of the survey footprint was covered
and the data were insufficiently complete 
to support a homogeneously calibrated source catalogue.
The goal of this paper is to present the next release 
that takes the coverage up to 92 percent of the survey area 
and includes a global photometric re-calibration.
In this work we
(i) explain the data reduction 
and quality control procedures that were applied,
(ii) describe and test the new photometric calibration, and 
(iii) detail the construction of the source catalogue
and demonstrate its use.

In \S\ref{sec:observations} we start by recapitulating the key points
of the survey observing strategy.
In \S\ref{sec:reduction} we describe the data reduction
and quality control procedures.
In \S\ref{sec:calibration} we explain the global re-calibration, in which
we draw upon the AAVSO Photometric All-Sky Survey (APASS)
and test our results against the Sloan Digital Sky Survey (SDSS).
In \S\ref{sec:catalogue} we explain how the source catalogue was compiled.
In \S\ref{sec:discussion} we discuss the properties of the catalogue
and in \S\ref{sec:demonstration} we demonstrate
the scientific exploitation of the colour/magnitude diagrams.
Finally, in \S\ref{sec:dataaccess} we discuss access
to the catalogue, an online library of reduced images and relevant source code,
The paper ends with conclusions in \S\ref{sec:conclusions} where we also outline
our future ambitions.

@@

The detailed properties of the IPHAS observing programme 
have been presented before 
by \citet{Drew2005} and \citet{Gonzalez-Solares2008}. 
To set the stage for this release, we provide a reminder of some key points.
IPHAS is an imaging survey of the Galactic Plane north of the celestial equator, 
from which photometry in Sloan $r, i$ 
is extracted along with narrowband H$\alpha$. 
It is carried out using the Wide Field Camera (WFC) 
on the 2.5-metre INT in La Palma. 
It is the first digital survey to offer comprehensive optical CCD photometry
of point sources in the Galactic Plane;
the footprint spans a box 
of roughly 180 by 10 degrees, 
covering Galactic latitudes $-5\degr < b < +5\degr$ 
and longitudes $30\degr < \ell < 215\degr$.

@@

\begin{figure*}
        \includegraphics[width=1\linewidth]{./plots/footprint/footprint_small.png}
        \caption{Survey area showing the footprints
        of all the quality-approved IPHAS fields
        which have been included in this data release.
        The area covered by each field has been coloured black
        with a semi-transparent opacity of 20\%,
        such that regions where fields overlap are darker.
        The IPHAS strategy is to observe each field twice
        with a small offset,
        and hence the vast majority of the area 
        is covered twice (dominant grey colour).
        There are small overlaps between all the neighbouring fields
        which can be seen as a honeycomb pattern
        of dark grey lines across the survey area.
        Regions with incomplete data are apparent as white gaps (no data) 
        or in light grey (denoting that only one offset is missing).
        The dark vertical strip near $\ell \sim$125\degr
        is an arbitrary consequence of the tiling pattern,
        which was populated starting from 0h in right ascension.}
        \label{fig:footprint}
\end{figure*}

Fig.~\ref{fig:footprint} shows the footprin
of the quality-approved observations included in this work. 
The fields which remain missing 
-- covering 8\% of the survey area --
are predominantly located towards the Galactic anti-centre 
at $\ell > 120\degr$.
Fields at these longitudes are mainly accessed from La Palma 
in the months of November-December,
which is when the La Palma weather and seeing conditions are often poor
forcing many (unsuccessful) repeat observations.
To enable the survey to be brought to completion, 
a decision was made recently to limit repeats in this area 
to individual fields requiring replacement,
i.e. fresh observations in all 3 filters may only be obtained 
at one of the two offset positions.  
The catalogue is structured such that it is clear 
where a contemporaneous observation of both halves of a field pair
is available.

@@

All raw IPHAS data were transferred
to the Cambridge Astronomical Survey Unit (CASU) 
for initial processing and archival.
The procedures used by CASU were originally devised
for the INT Wide Field imaging Survey \citep[WFS;][]{McMahon2001,Irwin2005},
which was a 200 deg$^2$ survey programme carried out 
between 1998 and 2003 after the WFC was commissioned.
Because IPHAS uses the same telescope and camera combination,
we have been able to benefit from the existing WFS pipeline.
A detailed description of the processing steps 
is found in \citet{Irwin2001}.
Its application to IPHAS has previously been described
by \citet{Drew2005} and \citet{Gonzalez-Solares2008},
and the source code is available
on-line\footnote{http://casu.ast.cam.ac.uk/surveys-projects/software-release}. 
In brief, the pipeline takes care of bias subtraction,
linearity correction, flat-fielding,
gain correction and de-fringing.

The reduced images are then stored in multi-extension FITS files
with a primary header describing the characteristics
(position, filter, exposure time, etc.) 
and four image extensions 
corresponding to each of the four CCDs.
Source detection and characterisation is then carried out 
using the \textsc{imcore} tool \citep{Irwin1985,Irwin1997}.
The flux of each source is measured using both
the peak pixel height
(i.e. a square 0\farcs 33$\times$0\farcs 33 aperture)
and a series of circular apertures of increasing diameter 
(1\farcs 2, 2\farcs 3, 3\farcs 3, 4\farcs 6 and 6\farcs 6).

@@

Finally, an astrometric solution is determined
based on the Two-Micron All Sky Survey (2MASS) point source catalogue \citep{Skrutskie2006},
which itself is calibrated 
in the International Celestial Reference System (ICRS).
A provisional photometric calibration is also provided 
based on the average zeropoint
determined from sets of standard stars observed within the same night.
Sources are classified morphologically
-- stellar, extended or noise --
based on the curve-of-growth determined
from measuring the source intensity in a series of growing apertures.
Finally, the resulting source detection tables are stored 
in multi-extension FITS files.

At the time of preparing DR2,
the CASU pipeline had processed
74\,195 IPHAS exposures 
in which a total of 1.9~billion \emph{candidate sources} were detected 
at the default detection level of 1.25\,$\sigma$.
This total inevitably includes spurious detections, artefacts and
duplicate detections;
in \S\ref{sec:catalogue} we will explain
how these have been removed or flagged in the final catalogue.
The pipelined data set -- comprising 2.5~terabyte of FITS files --
was then transferred to the University of Hertfordshire
for the purpose of transforming the raw
detection tables into a reliable source catalogue which is 
(i) quality-controlled,
(ii) homogeneously calibrated, and 
(iii) contains user-friendly columns and warning flags.
It is these post-processing steps which are explained next.

@@

Observing time for IPHAS is obtained
on a semester-by-semester basis
through the open time allocation committees 
of the Isaac Newton Group of telescopes.
The survey is allocated specific observing dates
rather than particular observing conditions.
In consequence, data were acquired
under a large range of atmospheric conditions.
Data taken under unsuitable conditions
have been flagged and removed
using seven quality criteria,
which ensure a reliable and homogeneous level of quality
across the data release:

\begin{figure}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=\textwidth]{./plots/depth_r.pdf} 
    \end{minipage}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=\textwidth]{./plots/depth_i.pdf} 
    \end{minipage}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=\textwidth]{./plots/depth_h.pdf} 
    \end{minipage}
    \caption{Distribution of the 5$\sigma$ limiting magnitude
             across all quality-approved survey fields
             for $r$ (top), $i$ (middle) and \ha\ (bottom).
             Fields with a limiting magnitude brighter than
             20th ($r$) or 19th (\ha, $i$) were rejected
             from the data release.
             The $r$-band depth is most sensitive 
             to the presence of the moon above the horizon: 
             this is the main reason for the wide, bi-modal character
             of its distribution.}
    \label{fig:depth}
\end{figure}

(1) \emph{Depth.} 
We discarded any exposures for which the $5\sigma$ limiting magnitude 
was worse than 20th magnitude in the $r$-band
or worse than 19th in $i$ or \ha. 
Such data were typically obtained during poor weather or full moon.
Most observations were significantly better than these limits.
Fig.~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields;
the mean depths and standard deviations are 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
The depth achieved depended
most strongly on the presence of the moon,
which was above the horizon during 62\% 
of the observations.
The great range in sky brightness this
produced is the main reason for the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depths attained in $i$ and \ha\ 
are less sensitive to moonlight, leading to
narrower magnitude limit distributions
(middle and bottom panels in Fig.~\ref{fig:depth}).
To a lesser extent, the wide spread in the $r$-band depth
is also explained by the shorter $r$-band exposure time
that was used in the first year of data-taking.
